\chapter*{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}

{\singlespace
This work would not be possible without the help and support of many of my colleagues, friends, and family.

First, I would like to express the deepest gratitude to my committee chair, Professor Ted Kim. I was immediately inspired by your pattern formation class. Somehow, I managed to maneuver into the study of computer graphics despite having little background experience. You were always very patient with me and managed to guide me through the various minefields of C++ programming and computer graphics algorithms. There were many moments of difficulty and frustration, but you helped me persevere in the end.

I am very grateful to Professor JoAnn Kuchera-Morin for serving on my committee. Your computer music class on Gamma helped inspire me to pursue the degree in Media Arts and Technology in the first place. We had many interesting and valuable discussions on the compositional process. 

My sincere appreciation goes to Professor Clarence Barlow for serving on my committee. Every single one of your music courses was a revelation to me, and I certainly would not be in the Media Arts and Technology department without your influence. 

My research in data compression owes a great debt to my collaborator Professor Pradeep Sen, who provided many insightful comments and ideas.

My thanks to my fellow classmates in Media Arts and Technology, who were a source of both technical and creative inspiration: Karl Yerkes, Sahar Sajadieh, Hannah Wolfe, Pablo Colapinto, Charlie Roberts, Michael Hetrick, Sterling Crispin, Yun Teng, Qiaodong Cui, Yuxiang Wang, and many others along the way.

My thanks to my friends nearby and far away, who provided much-needed diversion as well support in difficult times: Sarah Davis, Kyle Stewart, Chris Mendes, Maritza Fuljencio, Matt Guidry, Jason Garfield, Kyle Miller, Mike Morris, 
Jason Rothschild, Jamie Pommersheim, and Clara Bakker.

Finally, to my mom---thank you for supporting and believing in me, no matter what. To my partner Lauren, for many years of support, love, and encouragement. And lastly, to my cat, Professor Charles, who has been waiting long enough for the last member of the family to get a PhD.
}

