{
\singlespace
\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\chapter*{Vita of Aaron Demby Jones}
\markboth{Vita of Aaron Demby Jones}{Vita of Aaron Demby Jones}

\subsection*{Contact Information}
\vspace{.05in}
\begin{tabular}{@{}p{3.5in}p{4in}}
Media Arts and Technology Program & {\it Phone:}  (610) 334-1064\\             
University of California, Santa Barbara & {\it E-mail:} aaron.demby.jones@mat.ucsb.edu\\       
Santa Barbara, CA 93106-5080\\
\end{tabular}

\subsection*{Education}
  {\bf University of California, Santa Barbara} \hfill {\bf 2011--2017}\\
  {\em PhD Candidate, Media Arts and Technology Program} \hfill Santa Barbara,
  California\\[-3ex]

  {\bf University of Rochester}  \hfill {\bf 2009--2011}\\
  {\em BA, Mathematics} \hfill Rochester, New York\\[-3ex]
  
  {\bf Brown University}  \hfill {\bf 2005--2009}\\
  {\em BA, Music, Mathematics} \hfill Providence, Rhode Island\\[-3ex]

\subsection*{Honors and Awards} 
Chancellor's Fellowship, University of California, Santa Barbara \hfill {\bf 2011--2017}\\
Best Paper Award, Symposium on Computer Animation \hfill  {\bf 2016}\\
Muriel Hassenfeld Mann Premium, Brown University \hfill  {\bf 2009}\\
Buxtehude Premium, Brown University \hfill {\bf 2008}\\
Margery MacColl Award for Musical Excellence, Brown University  \hfill {\bf 2007}\\

\subsection*{Publications}
{\bfseries Jones, Aaron Demby,} Kuchera-Morin, JoAnn, and Kim, Theodore. Seeing and hearing the eigenvectors of a fluid. In {\itshape Proceedings of Bridges 2017: Mathematics, Art, Music, Architecture, Education, Culture.} Tessellations Publishing, 2017.

{\bfseries Jones, Aaron Demby,} Sen, Pradeep, and Kim, Theodore. Compressing fluid subspaces. In {\itshape Proceedings of the 2016 ACM SIGGRAPH/Eurographics Symposium on ComputerAnimation.} Eurographics Association, 2016.

\subsection*{Research Experience}

{\bf University of California, Santa Barbara} \hfill {\bf 2011--2017}\\
{\em Ph.D. Candidate} \hfill Santa Barbara, California\\
Sonification of subspace methods in computational fluid dynamics for computer graphics.

Advisor: Theodore Kim\\
Media Arts and Technology Program

{\bf Brown University} \hfill {\bf 2008--2009}\\
{\em Senior Honors Thesis} \hfill Providence, Rhode Island\\
Percussion Quartet

Advisor: Gerald Shapiro\\
Department of Music

\subsection*{Teaching Experience}
{\bf University of California, Santa Barbara}  \hfill {\bf Spring 2015}\\
{\em Teaching Assistant, Pattern Formation}\hfill  Santa Barbara, California\\

{\bf University of California, Santa Barbara} \hfill {\bf Fall 2014}\\
{\em Teaching Assistant, Music and Technology}\hfill  Santa Barbara, California\\

{\bf University of California, Santa Barbara} \hfill {\bf 2012--2013}\\
{\em Teaching Assistant, Music Appreciation}\hfill  Santa Barbara, California\\

{\bf Hampshire College} \hfill {\bf Summer 2014}\\
{\em Junior Faculty\\
Hampshire College Summer Studies in Mathematics} \hfill  Amherst, Massachusetts\\

{\bf Franklin and Marshall College} \hfill {\bf Summer 2012--2013, 2015--2017}\\
{\em Instructor, Number Theory\\
Johns Hopkins Center for Talented Youth} \hfill Lancaster, Pennsylvania\\

{\bf University of Rochester} \hfill {\bf 2010--2011}\\
{\em Teaching Assistant, Linear Algebra with Differential Equations}\hfill  Rochester, New York\\

{\bf University of Rochester} \hfill {\bf 2009--2010}\\
{\em Teaching Assistant, Calculus II}\hfill  Rochester, New York\\

\subsection*{Community Involvement}
{\bf Peer Review \hfill Fall 2016}\\
Reviewer for Eurographics 2017; Computer Graphics Forum}
